---
title: "The Most Expensive and Deadliest Weather Events in the United States, 1950-2011"
author: "apyle@github.com"
date: "2015-09-16"
output: 
  html_document:
    keep_md: true
---

<br />
<hr />

## Synopsis

Answer these two questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
1. Across the United States, which types of events have the greatest economic consequences?

<br />
<hr />

## Data

Actual dollars, not adjusted for inflation

Assumes no duplicate records
<br />
<hr />

## Data Processing

```{r setoptions, echo = TRUE}
# keep the chatter down so the write-up stays clean
options(warn = -1)

# make sure we have access to the knitr library for setup and housekeeping
# suppressWarnings should not be necessary since we have used the warn option
# but using it to reinforce the idea that we want to keep the write-up clean.
suppressWarnings(library(knitr))

# use data.table for ease of storage
suppressWarnings(library(data.table))

# use dplyr for data manipulation
suppressWarnings(library(dplyr))

# set knitr default option values, specifically echo, turning on the cache and
# setting its location, and specifying the figure directory
opts_chunk$set(echo = TRUE, 
               cache = TRUE, 
               cache.path = "cache/", 
               fig.path = "figure/")
```

<br />
<hr />

```{r}

# We only need the injury and fatality counts to answer the first question and the 
# property and crop damage for the second question. This will allow us to not load 
# the values we don't need, saving memory and processing time
col_classes <- c("NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", 
                 "character", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", 
                 "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", 
                 "numeric", "numeric", "numeric", "factor", "numeric", "factor", 
                 "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", "NULL", 
                 "numeric")
storm.file <- bzfile("repdata_data_StormData.csv.bz2")
storm.data <- data.table(read.csv(storm.file, colClasses = col_classes))

# most event entries do not have any reported fatalities, injuries, or damages. Filter those out
storm.subset <- storm.data[!(FATALITIES == 0 & INJURIES == 0 & PROPDMG == 0 & CROPDMG == 0), ]

# Just for sanity checks. Not really needed for analysis
my.factor <- as.factor(storm.data$EVTYPE)
summary(my.factor)
levels(my.factor)

```

There are 48 weather events categorized by NOAA. The input data has 985. In order 
to analyze the data and find the most dangerous and expensive events we much 
clean up the `EVTYPE` field in the input file. This applies to answering both 
questions.

```{r cache=FALSE}

# these are NOAA's weather events. Use this list to clean up the values found in the input file
event.list <- c(
              "Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", 
                "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", 
                "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", 
                "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Freezing Fog",
                "Frost/Freeze", "Funnel Cloud", "Hail", "Heat", 
                "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", 
                "Hurricane/Typhoon", "Ice Storm", "Lakeshore Flood", "Lake-Effect Snow",
                "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind",
                "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", 
                "Storm Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", 
                "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", 
                "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
event.list <- data.table(event.list)
event.list <- mutate(event.list, search = toupper(event.list))
setkey(event.list, search)

# add a variable in our data set to store the cleaned up version of the event type
storm.subset <- mutate(storm.subset, search = toupper(EVTYPE))
setkey(storm.subset, search)

# Match the input data with the NOAA list
storm.set1 <- left_join(storm.subset, event.list, by = c("search" = "search"))

# how many are matched with good EVTYPE values?
records.total <- count(storm.set1)
records.unmatched <- sum(is.na(storm.set1$event.list))
records.matched <- sum(!is.na(storm.set1$event.list))
records.ratio <- round(records.matched / records.total, 3)

```

There are a total of `r records.total` records. Of these records, there are `r records.matched` 
which correctly coded as NOAA's recognized weather events. This is a ratio of `r records.ratio * 100`% 
that were well coded. The remaining unmatched records represent a significant number of fatalities, 
injuries, and damage that could change the results of the most dangerous and expensive weather events. 
In order to keep this miscoding from under-reporting the results, we'll clean up as much of the data 
as we can.

Some of the data records include multiple weather events such as Thunderstorm Wind and Lightning or 
Heavy Snow and High Wind. In these cases we will categorize the record under the first weather event. 
In these two examples, the data record would be recorded as Thunderstorm Wind and Heavy Snow 
respectively and the data would not appear as Lightning or Heavy Snow. This will prevent fatalities, 
injuries, and damages from appearing multiple times which would skew the total. The second, or third, 
events listed will be under reported in these cases. Another approach would be to divide the damages 
evenly between the weather events. However, this proves to be very time consuming for a marginal 
refinement of the results.

```{r cache=FALSE}
# most of the records were coded with the correct EVTYPE, i.e., the ones in the NOAA weather event list.
# Now we need to clean up as many of the incorrectly coded EVTYPEs

# there are a large number of Thunderstorm Winds that are not coded correctly. 
is.tstm <- substr(storm.set1$search, 1, 9) == "TSTM WIND"
storm.set1$event.list[is.tstm] <- c("Thunderstorm Wind")
# some are spelled out but also include plurals or suffixes
is.tstm <- substr(storm.set1$search, 1, 17) == "THUNDERSTORM WIND"
storm.set1$event.list[is.tstm] <- c("Thunderstorm Wind")
# some are spelled out with extra plurals
is.tstm <- substr(storm.set1$search, 1, 19) == "THUNDERSTORMS WINDS"
storm.set1$event.list[is.tstm] <- c("Thunderstorm Wind")

# Marine Thunderstorm Winds are also coded as TSTM
is.marine <- grep("MARINE TSTM", storm.set1$search)
storm.set1$event.list[is.marine] <- c("Marine Thunderstorm Wind")

# Flooding is categorized in many ways
is.flood <- grep("URBAN/SML STREAM FLD", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("FLOOD/FLASH FLOOD", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("RIVER FLOOD", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("URBAN FLOODING", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("URBAN FLOOD", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("FLOODING", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("COASTAL FLOODING", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("RIVER FLOODING", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("FLOOD/RAIN/WINDS", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("RIVER FLOODING", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")
is.flood <- grep("FLOODS", storm.set1$search)
storm.set1$event.list[is.flood] <- c("Flood")

# there are a large number of Flash Floods that are not coded correctly. 
is.tstm <- substr(storm.set1$search, 1, 11) == "FLASH FLOOD"
storm.set1$event.list[is.tstm] <- c("Flash Flood")

# there are a large number of Hail events that are not coded correctly. 
is.tstm <- substr(storm.set1$search, 1, 4) == "HAIL"
storm.set1$event.list[is.tstm] <- c("Hail")

# Prefer Excessive Heat when there is a descriptor
is.heat <- substr(storm.set1$search, 1, 9) == "HEAT WAVE"
storm.set1$event.list[is.heat] <- c("Excessive Heat")
is.heat <- substr(storm.set1$search, 1, 12) == "EXTREME HEAT"
storm.set1$event.list[is.heat] <- c("Excessive Heat")

# Heat often has descriptors that make it hard to code correctly
is.cold <- substr(storm.set1$search, 1, 12) == "EXTREME COLD"
storm.set1$event.list[is.cold] <- c("Extreme Cold/Wind Chill")

# several items are listed as plural while the standard is singular
is.plural <- grep("HIGH WINDS", storm.set1$search)
storm.set1$event.list[is.plural] <- c("High Wind")

is.plural <- grep("RIP CURRENTS", storm.set1$search)
storm.set1$event.list[is.plural] <- c("Rip Current")


# Hurricanes & Typhons are very damaging. Make sure we include these rare but important events in the results
# These grep statements will include the entries already matched but does not cause a problem
is.ht <- grep("HURRICANE", storm.set1$search)
storm.set1$event.list[is.ht] <- c("Hurricane/Typhoon")
is.ht <- grep("TYPHOON", storm.set1$search)
storm.set1$event.list[is.ht] <- c("Hurricane/Typhoon")

# Tropical Storms are too
is.ts <- grep("TROPICAL STORM", storm.set1$search)
storm.set1$event.list[is.ts] <- c("Tropical Storm")

# clean up some Tornados
is.tornado <- substr(storm.set1$search, 1, 7) == "TORNADO"
storm.set1$event.list[is.tornado] <- c("Tornado")

# Storm Tide comes across as Storm Surge
is.st <- substr(storm.set1$search, 1, 11) == "STORM SURGE"
storm.set1$event.list[is.st] <- c("Storm Tide")

#  Heavy rain has additional terms
is.rain <- substr(storm.set1$search, 1, 10) == "HEAVY RAIN"
storm.set1$event.list[is.st] <- c("Heavy Rain")

# after our clean-up how many are matched with good EVTYPE values?
records.total <- count(storm.set1)
records.unmatched <- sum(is.na(storm.set1$event.list))
records.matched <- sum(!is.na(storm.set1$event.list))
records.ratio <- round(records.matched / records.total, 3)

```

After cleaning up the data there are still a total of `r records.total` records of which we now have matched `r records.matched` for a total of `r records.ratio * 100`%.

```{r cache=FALSE}
# the input data records property and crop damages in two fields, one for a numerical amount 
# and another for the magnitude of damages. We'll need to get the actual number for our analysis

# adjust property damage costs based on the PROPGMGEXP
storm.set1 <- storm.set1 %>%
        mutate(prop.total = PROPDMG) %>%
        mutate(prop.total = ifelse(PROPDMGEXP == "K", prop.total * 1000, prop.total)) %>%
        mutate(prop.total = ifelse(PROPDMGEXP == "M", prop.total * 1000000, prop.total)) %>%
        mutate(prop.total = ifelse(PROPDMGEXP == "B", prop.total * 1000000000, prop.total))

# and do the same for crop damage costs
storm.set1 <- storm.set1 %>%
        mutate(crop.total = CROPDMG) %>%
        mutate(crop.total = ifelse(CROPDMGEXP == "K", crop.total * 1000, crop.total)) %>%
        mutate(crop.total = ifelse(CROPDMGEXP == "M", crop.total * 1000000, crop.total)) %>%
        mutate(crop.total = ifelse(CROPDMGEXP == "B", crop.total * 1000000000, crop.total))

```

## Results

```{r dangerous, cache=FALSE}
# assemble the reporting data structure from the cleaned up data

# first, assemble the reporting data for dangerous weather events
by_event <- group_by(storm.set1, event.list)
dangerous <- summarize(by_event, sum(FATALITIES), sum(INJURIES))
setnames(dangerous, 1:3, c("Weather Event", "Deaths", "Injuries"))
dangerous <- dangerous %>% 
        #na.omit() %>% 
        mutate(Total = Deaths + Injuries) %>% 
        mutate(rev.sort = 100000 - Total)

setkey(dangerous, rev.sort)

dangerous[1:47, ]
#summarize(dangerous$Total)
```
<br />
<hr />

```{r costly, cache=FALSE}
# assemble the reporting data for weather event costs
by_event <- group_by(storm.set1, event.list)
costs <- summarize(by_event, sum(prop.total), sum(crop.total))
setnames(costs, 1:3, c("Weather Event", "Property", "Crop"))
costs <- costs %>% 
        #na.omit() %>% 
        mutate(Total = Property + Crop) %>% 
        mutate(rev.sort = 100000000 - Total)


setkey(costs, rev.sort)

#na.omit(costs[1:47, ])
costs[1:50, ]
```

## Conclusion

